Tracking p-adic precision: examples

### 2014-08-04 examples.tex
Starting a new paper: examples.tex. Until now:  - I wrote a technical result which may be useful (hopefully)    to bound many growing functions Lambda(f), and  - I copy-pasted the removed appendix of our latest paper.

%\documentclass{sig-alternate}
\documentclass{lms}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb}
\usepackage{amsrefs}
\usepackage{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\renewcommand*{\theHsection}{\thesection}

@@

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
%\theoremstyle{definition}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

@@

@@

\title{Tracking $p$-adic precision}
\author{Xavier Caruso, David Roe and Tristan Vaccon}
\date\today

\maketitle
\begin{abstract}
We present a new method to propagate $p$-adic precision in computations, which also
applies to other ultrametric fields.
We illustrate it with many examples and give a toy application to the
stable computation of the SOMOS 4 sequence.
\end{abstract}

\setcounter{tocdepth}{1}
\tableofcontents

@@

The last two decades have seen a rise in the popularity of $p$-adic methods in
computational algebra.  For example,
\begin{itemize}
\item Bostan et al. \cite{boston-gonzalez-perdry-schost:05a} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
\item Gaudry et al. \cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
\item Kedlaya \cite{kedlaya:01a}, Lauder \cite{lauder:04a} and many followers used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
\item Lercier and Sirvent \cite{lercier-sirvent:08a} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
\end{itemize}
Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
that apply across many different algorithms.

Two sources of inspiration arise when studying $p$-adic algorithms.
The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
homomorphism, given unknown elements in two such balls we can
locate the balls in which their sum and product lie.  Working on a computer
we must find a way to write elements using only a finite amount of data.
By lumping elements together into these balls of radius $p^{-n}$, we may
model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
all $p$-adic elements have constant \emph{absolute precision} $n$.

The second source draws upon parallels between $\Qp$ and $\R$.  Both
occur as completions of $\Q$ and we represent elements of both in terms
of a set of distinguished rational numbers.
In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
\[
\Big\vert \frac{x \circledast y}{x \ast y} - 1 \Big\vert \le 2^{-h}
\]
for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
\[
\Big\vert \frac{x \circledast y}{x \ast y} - 1 \Big\vert_p \le p^{-h}.
\]
When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.

In both of these models, precision (absolute or relative) is constant across all elements.
Since some operations lose precision, it can be useful to attach a precision to each element.
Over the reals, such interval arithmetic is unwieldy, since arithmetic operations always increase
the lengths of the inputs.  As a consequence, most computations in the real numbers rely on
statistical cancelation and external estimates of precision loss, rather than
attempting to track known precision at each step.  This tendency is strengthened
by the ubiquity of floating point arithmetic in scientific applications, where Gaussian
distributions are more common than intervals anyway.

In the $p$-adic world, precision tracking using intervals is much more feasible.
Even a long sequence of operations with such elements may not sacrifice any precision.  Intervals
allow number theorists to provably determine a result modulo a given power of $p$,
and the Gaussian distributions of measurement error over $\R$ have no direct analogue
over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
in implementations of $p$-adic numbers.  The mathematical software packages
Sage \cite{sage}, PARI \cite{pari} and Magma \cite{magma} all include $p$-adic elements
that track precision in this way.

The approach of propagating precision with each arithmetic operation works well, but does
sometimes underestimate the known precision of a result, as we will discuss in Section
\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
implementations of polynomials, vector spaces, matrices and power series.  The practice
of storing the precision within each entry is not flexible enough for all applications.
Sometimes only a rough accounting of precision is needed, in which case storing
and computing the precision of each entry in a large matrix needlessly consumes
space and time.  Conversely, recording the precision of each entry does not allow
a constraint such as specifying the precision of $f(0)$, $f(1)$ and $f(2)$ for a
quadratic polynomial $f$.

For a vector space $V$ over $\Qp$, we propose that the fundamental object used to
store the precision of an element should be a $\Zp$-lattice $H \subset V$.  By using
general lattices one can eliminate needless loss of precision.  Moreover, specifying
the precision of each entry or recording a fixed precision for all entries can both
be interpreted in terms of lattices.  In Section \ref{sec:prec-proposal} we detail our
proposal for how to represent the precision of an element of a vector space.

In Section \ref{sec:mainlemma}, we develop the mathematical background on which our proposal is based.
The most notable result of this section is Lemma \ref{lem:main} which describes how lattices transform under non-linear maps and allows us to propagate
precision using differentials.  More specifically, it describes a class of first order lattices,
whose image under a map of Banach spaces is obtained by applying the
differential of that map.  In Section \ref{ssec:locanalytic} we make the conditions
of Lemma \ref{lem:main} more explicit in the case of locally analytic functions.

In Section \ref{sec:tracking} we propose methods for tracking precision in practice.
Section \ref{ssec:opt-tracking} includes a discussion of two models of precision tracking:
one-pass tracking, where the precision lattice is propagated at each step of the algorithm,
and two-pass tracking, where an initial pass computing rough approximations is used
in computing the precision lattices.  We introduce precision types in Section \ref{ssec:types},
which allow a tradeoff between flexibility, space and time in computing with precision.
In Section \ref{ssec:SOMOS-solution}, we give
an application of these ideas to an algorithm for computing terms of the SOMOS sequence.

In Appendix \ref{sec:manifold}, we extend the results of Section \ref{sec:mainlemma}
to $p$-adic manifolds, describing how to specify precisions for points on elliptic curves
and Grassmannians.  Finally, Appendix \ref{sec:matrices} describes how to compute
the derivative of some common operations on matrices as an example, with an eye
toward applying Lemma \ref{lem:main}.

\section{Precision proposals} \label{sec:prec-proposal}

\subsection{Problems in precision}
\label{ssec:stepbystep}

The usual way to track $p$-adic precision consists in replacing $p$-adic 
numbers by approximate elements of the form $a + O(p^N)$ and performing 
all usual arithmetical operations on these approximations.
%Recall that 
%an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if it 
%is well defined up to adding an unknown element of $p^N\Zp$.
We offer below three examples that illustrate cases where this way to 
track precision does not yield optimal results.

\subsubsection*{A linear map.}

Consider the function $f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem
of computing $f \circ f(a + O(p^n), b + O(p^m))$.  Applying $f$ twice, computing precision
with each step, yields $\left(2a + O(p^{\min(m,n)}), 2b + O(p^{\min(m, n)})\right)$.
On the other hand, $f \circ f(x,y) = (2x, 2y)$, so one may compute the result more
accurately as $(2a + O(p^n), 2b + O(p^m))$, with even more precision when $p=2$.

\subsubsection*{SOMOS 4.}

The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
\[
u_n = \frac{u_{n-3} u_{n-1} + u_{n-2}^2}{u_{n-4}}.
\]
We shall consider the case where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
Let us first examine how the absolute precision of $u_n$ varies with $n$
if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
The computation of $u_n$ involves a division by $u_{n-4}$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_{n-4})}$.  Hence the
step-by-step computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

On the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}: for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
From the latter formula and our assumption that $u_0$, $u_1$, 
$u_2$ and $u_3$ are units known up to precision $O(p^N)$, it follows that all $u_n$'s
are known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm used to compute $u_n$.

\begin{rmk}
From the above discussion, one can easily derive a numerically stable 
algorithm that computes the SOMOS 4 sequence: 
\begin{enumerate}
\item compute the polynomials $P_n$ using the recurrence in 
the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$
\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
\end{enumerate}
However, computing the $P_n$'s is very time-consuming since it requires 
division in a polynomial ring with $4$ variables and the
size of the coefficients of $P_n$ explodes as $n$ grows.

In Section \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
the SOMOS 4 sequence which turns out to be, at the same time, efficient 
and numerically stable.
\end{rmk}

\subsubsection*{LU factorization.}

Let us first recall that a square matrix $M$
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ are lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{sec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the Cramer-type formulae
in the second algorithm yield a substantially longer running time than the first.
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

\subsection{Lattices}

In order to make our proposals for tracking precision clear, we need
some definitions from ultrametric analysis.  See 
Schneider \cite{schneider:11a} for a more complete exposition.

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric (\emph{i.e.}
$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a computable dense subring $R \subset K$ \cite{rabin:60a}.
This assumption holds for $K = \Qp$ and $K=\Fp\lp t \rp$ by setting $R = \Z[\frac1p]$ or $R = \Q$
in the case of $\Qp$ and $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$ in the case of $\Fp\lp t \rp$.

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
We underline that any lattice $H$ in $E$ is also closed since its 
complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
are all open.
For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
$$B_E(r) = \{ x \in E : \Vert x \Vert \le r \}, 
\quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.  We will also set $B_E(\infty) = B^-_E(\infty) = E$.

Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$ (according to the filter of cofinite subsets), and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.

Given a basis $(x_i)_{i \in I}$ and a sequence $(r_i)_{i \in I}$ with $r_i \in \R_{>0}$, the sets
\begin{align*}
B_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert \le r_i \Big\}, \\
B^-_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert < r_i \Big\}
\end{align*}
are lattices precisely when the $r_i$ are bounded.  If we have equipped $E$ with a distinguished basis
then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.

\subsubsection*{Approximate elements.}

Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
\begin{deftn}
\label{def:approximate}
\begin{itemize}
\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
\begin{equation} \label{eq:exact_elt}
x = \sum_{j \in J} \alpha_j x_j.
\end{equation}

\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
and $H$ is a lattice in $E$.
\end{itemize}
\end{deftn}

The pair $(x, H)$ represents an undetermined element of the coset
$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.

\subsubsection*{Lattices and computers.}

>>>>

\documentclass{sig-alternate}
%\documentclass{lms}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb}
\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\renewcommand*{\theHsection}{\thesection}

@@

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
%\theoremstyle{definition}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

@@

\newcommand{\id}{\textrm{id}}
\newcommand{\Epi}{\textrm{Epi}}

@@

\def\binom#1#2{\Big(\begin{array}{cc} #1 \\ #2 \end{array}\Big)}

\title{Tracking p-adic precision: examples}

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
\alignauthor David Roe \\
  \affaddr{University of Calgary}\\
  \email{\normalsize \textsf{roed.math@gmail.com}}
}

\maketitle

\begin{abstract}
\end{abstract}

\vspace{1mm}
 \noindent
 {\bf Categories and Subject Descriptors:} \\
\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
  Manipulation -- \emph{Algebraic Algorithms}

 \vspace{1mm}
 \noindent
 {\bf General Terms:} Algorithms, Theory

 \vspace{1mm}
 \noindent
 {\bf Keywords:} 
\medskip

@@

\section{Preliminaries}

In the next section, we will see that the derivative at some point $x$ 
of many standard operations, modelised by a function $f$, has a simple 
expression in term of $x$ and $f(x)$. In other terms, such functions $f$
satisfy:
\begin{equation}
\label{eq:diffequa}
f' = g \circ (f, \id)
\end{equation}
where $g$ is a given --- and hopefully rather simple --- function. 
The aim of these preliminaries is to study the differential equation
\eqref{eq:diffequa} (or, actually, a slight generalization of it) and
to derive from it some bounds on the growing function $\Lambda(f)$ 
defined in \cite{padicprec}.

\begin{comment}

\subsection{Legendre transform of a compositum}

For $i \in \{1,2\}$, let $\varphi : \R \to \R \cap \{+\infty\}$ denote a 
convex function and $\varphi_i^\star$ be its Legendre transform defined 
by:
\begin{equation}
\label{eq:defLegendre}
\varphi_i^\star(u) = \sup_{v \in \R} \varphi(v) - uv.
\end{equation}
Set $\varphi = \varphi_1 \circ \varphi_2$. The aim of this subsection is 
to relate $\varphi^\star$ with $\varphi_1^\star$ and $\varphi_2^\star$.

\begin{lem}
\label{lem:Legendrecomp}
We assume that $\varphi_1$ and $\varphi_2$ are nondecreasing and that 
their slopes are all in $\N$. Then, for all nonnegative interger $s$:
$$\varphi^\star(s) =
\inf_{\ell, (n_i)} \varphi_1^\star(\ell) + \varphi_2^\star(n_1) + 
\cdots + \varphi_2^\star(n_\ell)$$
where the infimum is taken over all finite sequences of nonnegative
integers (of variable length) $n_1, \ldots, n_\ell$ with $n_1 + \cdots +
n_\ell = s$.
\end{lem}

\begin{proof}
Let $n_1, \ldots, n_\ell$ be a sequence as above. Taking the supremum 
in $v$ in the relation:
$$\varphi(v) - s v =
\varphi_1(\varphi_2(v)) - \ell \varphi_2(v) + \sum_{i=1}^\ell
(\varphi_2(v) - n_i v)$$
we get:
\begin{equation}
\label{eq:inegvarphi}
\varphi^\star(s) \leq \varphi_1^\star(\ell) + 
\varphi_2^\star(n_1) + \cdots + \varphi_2^\star(n_\ell).
\end{equation}
It remains then to prove that one can choose $\ell$ and $n_1, \ldots,
n_\ell$ such that the inequality above is actually an equality.

Let $x$ be the unique value such that the left (resp. right) derivative 
of $\varphi$ at $x$ is less than (resp. greater than) or equal to $s$. 
Let $\ell^-$ (resp. $\ell^+$) denote the left (resp. right) derivative 
of $\varphi_1$ at $\varphi_2(x)$ and $n^-$ (resp. $n^+$) denote the left 
(resp. right) derivative of $\varphi_2$ at $x$. On the one hand, $\ell^- 
n^- \leq s \leq \ell^+ n^+$ since $\ell^- n^-$ (resp. $\ell^+ n^+$) 
equals the left (resp. right) derivative of $\varphi$ at $x$). On the 
other hand:
$$\begin{array}{ll}
\varphi_1^\star(u) = \varphi(x) - u \varphi_2(x) & 
\text{for all } u \in [\ell^-,\, \ell^+] \\
\varphi_2^\star(u) = \varphi_2(x) - u x & 
\text{for all } u \in [n^-,\, n^+] \\
\varphi^\star(u) = \varphi(x) - u x & 
\text{for all } u \in [\ell^- n^-,\, \ell^+ n^+] \\
\end{array}$$
Therefore, the inequality \eqref{eq:inegvarphi} is an equality for all 
integer $\ell \in [\ell^-, \ell^+]$ and all sequences of integers 
$(n_i)_{1 \leq i \leq \ell}$ with $n_i \in [n^-, n^+]$ for all $i$ and 
$n_1 + \cdots + n_\ell = s$. It is then enough to prove that such a 
couple $(\ell, (n_i))$ exists but it is clear.
\end{proof}

\end{comment}

\subsection{Composite of locally analytic functions}

Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces
$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$ and
consider two locally analytic functions $f : U \to V$ and $g : V \to W$ 
two locally analytic functions around $0$. The composition $h = g \circ
f$ is then locally analytic around $0$ as well. Let us write the
analytic expansion of $f$, $g$ and $h$ as follows:
$$f = \sum_{n \geq 0} f_n, \quad 
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n, \quad$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The aim of
this subsection is to prove the following proposition.

\begin{prop}
\label{prop:boundhr}
With the above notations, we have for all nonnegative integer $r$:
$$\Vert h_r \Vert \leq \sup_{m, (n_i)}
  \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert$$
where the supremum is taken over all couples $(m, (n_i))$ where $m$
is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
\end{prop}

During the proof, multinomial coefficients will play an important role. 
Given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
k_1 + \cdots + k_\ell$, it is defined by:
$$\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
  \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.$$
We recall that they are always positive integers.
We now expand $g \circ f$ as follows:
\begin{align}
g \circ f & = \sum_{m \geq 0} g_m \Big(\sum_{n \geq 1} f_n\Big) \nonumber \\
& = \sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell}, \ldots, f_{n_\ell})
\label{eq:expansiongf}
\end{align}
where the latest sum runs over:

\noindent
(a) all finite sequences $(k_i)$ of positive integers whose length 
(resp. sum) is denoted by $\ell$ (resp. $m$), and

\noindent
(b) all finite sequences $(n_i)$ of positive integers of length
$\ell$.

\noindent
Moreover, in the argument of $G_m$, the variable $f_{n_i}$ 
is repeated $k_i$ times.

The degree of $G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell},
\ldots, f_{n_\ell})$ is $r = k_1 n_1 + \ldots + k_\ell n_\ell$ and 
then contributes to $h_r$. As a consequence $h_r$ is equal to 
\eqref{eq:expansiongf} where the sum is restricted to sequences
$(k_i)$, $(n_i)$ such that $k_1 n_1 + \ldots + k_\ell n_\ell = r$.
Proposition \ref{prop:boundhr} now follows from the next lemma.

\begin{lem}
Let $\mathcal E$ be a $K$-vector space. Let $\varphi : \mathcal E^n \to 
K$ be a symmetric $s$-linear form and $\psi: E \to K$ defined by 
$\psi(x) = \varphi(x, x, \ldots, x)$.
Given $k_1, \ldots, k_\ell$ some positive integers whose sum is $m$ and 
$x_1, \ldots, x_\ell$ some elements of $\mathcal E$, we have:
$$\begin{array}{l}
\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
\varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
x_\ell) \Big\Vert  \\
\hspace{4.3cm} \leq \Vert \psi \Vert \cdot \Vert x_1 \Vert^{k_1} \cdots
 \Vert x_\ell \Vert^{k_\ell}
\end{array}$$
where, in LHS, the variable $x_i$ is repeated $k_i$ times.
\end{lem}

\begin{proof}
It is enough to prove that:
\begin{equation}
\label{eq:normpolar}
\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
\varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
x_\ell) \Big\Vert \leq \Vert \psi \Vert
\end{equation}
provided that all the $x_i$'s have norm at most $1$. We proceed by 
induction on 
$\ell$. If $\ell = 1$, Eq.~\eqref{eq:normpolar} follows directly from
the definition of $\Vert \psi \Vert$. 
We now pick $(\ell+1)$ integers $k_1, \ldots, k_{\ell+1}$ whose sum 
equals $m$ together with $(\ell+1)$ elements $x_1, \ldots, x_{\ell+1}$
lying in the unit ball of $\mathcal E$.
We also consider a new variable $\lambda$ varying in
$\O_K$. We set $x'_i = x_i$, $k'_i = k_i$ when $i < \ell$ and $x'_\ell 
= x_\ell + \lambda x_{\ell+1}$, $k_\ell = k_\ell + k_{\ell+1}$. By the 
induction hypothesis, we know that the inequality:
$$\Big\Vert \binom s {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
\varphi(x'_1, \ldots, x'_1, \ldots, x'_\ell, \ldots, x'_\ell) \Big\Vert
\leq \Vert \psi \Vert$$
holds for all $\lambda \in K$. Furthermore, LHS of the above inequality
is a polynomial $P(\lambda)$ of degree $k'_\ell$ whose coefficient in 
$\lambda^j$ is:
$$\binom s {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
\binom {k'_\ell} {j} \cdot
\varphi(\underline x_j) = 
\binom s {k_1 \,\, \cdots \,\, k_{\ell-1} \,\, j} \cdot
\varphi(\underline x_j)$$
with
$$\underline x_j = (x_1, \ldots, x_1, \ldots, x_{\ell+1}, \ldots, 
x_{\ell+1})$$
where $x_i$ is repeated $k_i$ times if $i < \ell$ and $x_\ell$ 
(resp. $x_{\ell+1}$) is repeated $j$ times (resp. $k'_\ell - j$ times).
Since $\Vert P(\lambda) \Vert \leq \Vert \psi \Vert$ for all $\lambda$ in 
the unit ball, the norm of all its coefficients must be at most $\Vert \psi
\Vert$ as well. Looking at the coefficient in $\lambda^{k_i}$, we find
Eq.~\eqref{eq:normpolar} for the families $(k_1, \ldots, k_{\ell+1})$ 
and $(x_1, \ldots, x_{\ell+1})$. The induction follows.
\end{proof}

\subsection{Bounding a growing function}
\label{ssec:boundLambdaf}

We now study a slight generalization of the differential equation 
\eqref{eq:diffequa}, which is:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a $K$-Banach space $G$. Consequently, $g$
goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
all continuous linear applications $E \to F$.
In what follows, we always assume that $V$ and $W$ contains the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are armless 
because (1)~we can always shift $f$ and $h$ (and $g$ accordingly) so 
that they both vanish at $0$ and (2)~in order to apply the theory of 
\cite{padicprec}, the derivative $f'(0)$ needs to be surjective and 
therefore \emph{a fortiori} nonzero.

Let us write the analytic expansion of $f$, $g$ and $h$ as follows: 
$$f = \sum_{n \geq 0} f_n, \quad
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
Following \cite{padicprec}, we define $\NP(f) : \R \to \R \cup 
\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
\log \Vert f_n \Vert$ for all integer $n$ and denote by $\Lambda(f)$ its 
Legendre transform. We define $\NP(g)$, $\Lambda(g)$, $\NP(h)$ and
$\Lambda(h)$ similarly.

We now assume that \textbf{$K$ has caracteristic $0$} and we consider a 
nonnegative real number $\alpha$ such that $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all positive integer $n$. A suitable value for 
$\alpha$ is
$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$
where $p$ denotes the characteristic of the residue field and
where, by convention, the above expression equals $0$ if $p = 0$.

\begin{prop}
\label{prop:boundLambdaf}
We keep the above notations. If the couple $(a,b)$ satisfies:
\begin{equation}
\label{eq:condAB}
b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)
\end{equation}
then $b \geq \Lambda(f)(a - \alpha)$.
\end{prop}

\begin{proof}
Assuming that $F_n$ is symmetric (which is armless in characteristic 
$0$), we get $f' = \sum_{n \geq 0} f'_n$ where
$$f'_n : U \to \mathcal L(E,F), \quad
x \mapsto \big(h \mapsto n \cdot F_n(h, x, x, \ldots, x)\big).$$
In particular, taking $h = x$, we find:
\begin{equation}
\label{eq:normderivative}
\Vert f'_n \Vert \geq \Vert n f_n \Vert = 
\Vert n \Vert \cdot \Vert f_n \Vert.
\end{equation}
Combining this with Proposition \ref{prop:boundhr}, we get, for all
nonnegative integer $r$:
$$\Vert (r+1) f_{r+1} \Vert \leq
  \sup_{m, (n_i)} \Vert g_m \Vert \cdot 
  \prod_{i=1}^m \max(\Vert f_{n_i} \Vert, \Vert h_{n_i} \Vert)$$
where the supremum runs over all couples $(m, (n_i))$ where $m$
is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
We set $u_r = \Vert r! f_r \Vert$. Multiplying the above inequality by
$\Vert r! \Vert$, we obtain:
\begin{equation}
\label{eq:boundurrec}
u_{r+1} \leq
  \sup_{m, (n_i)} \Vert g_m \Vert \cdot 
  \prod_{i=1}^m \max(u_{n_i}, \Vert n_i! h_{n_i} \Vert)
\end{equation}
since the multinomial coefficient $\binom r {\!n_1 \, \cdots \, n_m\!}$
is an integer and hence as norm at most $1$.
